Approximation as Differentiation

### 2025-06-09 main/approx-as-tangents.tex
Gentler notation of syntactic identifiers.

\begin{example}[Conditionally Multiplicative Functions]
  To see the effect of requiring preservation of bounded meets,
  consider several ways of defining disjunction on the lifted booleans
  $\mathbb{B}_\bot$. Two functions that are cm are the strict and
  left-short-circuiting ORs\footnote{The clauses in these examples are
    shorthand for the graph of the function. They are not to be
    understood as pattern matching clauses in a language like Haskell,
    where it is not possible to match on $\bot$.}:
  \begin{displaymath}
    \begin{array}[t]{l@{(}l@{,~}l@{)~}c@{~}l}
      \mathrm{strictOr}&\mathsf{tt}&\mathsf{tt}&=&\mathsf{tt} \\
      \mathrm{strictOr}&\mathsf{tt}&\mathsf{ff}&=&\mathsf{tt} \\
      \mathrm{strictOr}&\mathsf{ff}&\mathsf{tt}&=&\mathsf{tt} \\
      \mathrm{strictOr}&\mathsf{ff}&\mathsf{ff}&=&\mathsf{ff} \\
      \mathrm{strictOr}&\bot&\_&=&\bot \\
      \mathrm{strictOr}&\_&\bot&=&\bot \\
    \end{array}
    \qquad
    \begin{array}[t]{l@{(}l@{,~}l@{)~}c@{~}l}
      \mathrm{shortCircuitOR}&\mathsf{tt}&\_&=&\mathsf{tt} \\
      \mathsf{shortCircuitOR}&\mathsf{ff}&x&=&x \\
      \mathsf{shortCircuitOR}&\bot&\_& =& \bot
    \end{array}
    \qquad
    \raisebox{-15ex}{
      \begin{tikzpicture}
        \node (top) at (0,0) {$(\mathsf{tt},\mathsf{ff})$};
        \node [below of=top,xshift=-0.5cm] (oi) {$(\bot,\mathsf{ff})$};
        \node [below of=top,xshift=0.5cm] (io) {$(\mathsf{tt},\bot)$};
        \node [below of=oi,xshift=0.5cm] (oo) {$(\bot,\bot)$};
        \draw (top) -- (oi);
        \draw (top) -- (io);
        \draw (oi) -- (oo);
        \draw (io) -- (oo);
      \end{tikzpicture}
    }
  \end{displaymath}
  In the poset $\mathbb{B}_\bot^2$, a typical poset of approximations of a fully defined element is shown to the right. For $\mathrm{strictOr}$, any approximation that isn't the fully defined input is mapped to $\bot$, while $\mathrm{shortCircuitOr}$ maps the partially defined $(\mathsf{tt},\bot)$ to $\mathsf{tt}$. Thus, even though these functions operate identically on fully defined inputs, they differ in their \emph{derivatives} on partially defined input, exposing how . That they are cm can be checked by examining their restrictions' behaviour. If we take the approximations $(\bot,\mathsf{ff})$ and $(\mathsf{tt},\bot)$, then their meet is $(\bot,\bot)$; $\mathrm{strictOr}$ maps all three elements to $\bot$, so is cm here since $\bot \wedge \bot = \bot$; and $\mathrm{shortCircuitOR}$ has $(\bot,\mathsf{ff}) \mapsto \bot$ and $(\mathsf{tt},\bot) \mapsto \mathsf{tt}$, the meet of which is $\bot = \mathrm{shortCircuitOR}(\bot,\bot)$. Other combinations can be checked similarly.
\end{example}

>>>>

\begin{example}[Conditionally Multiplicative Functions]
\label{ex:strict-short-circuit}
  To see the effect of requiring preservation of bounded meets,
  consider several ways of defining disjunction on the lifted booleans
  $\mathbb{B}_\bot$. Two functions that are cm are the strict and
  left-short-circuiting ORs\footnote{The clauses in these examples are
    shorthand for the graph of the function. They are not to be
    understood as pattern matching clauses in a language like Haskell,
    where it is not possible to match on $\bot$.}:
  \begin{displaymath}
    \begin{array}[t]{l@{(}l@{,~}l@{)~}c@{~}l}
      \mathrm{strictOr}&\mathsf{tt}&\mathsf{tt}&=&\mathsf{tt} \\
      \mathrm{strictOr}&\mathsf{tt}&\mathsf{ff}&=&\mathsf{tt} \\
      \mathrm{strictOr}&\mathsf{ff}&\mathsf{tt}&=&\mathsf{tt} \\
      \mathrm{strictOr}&\mathsf{ff}&\mathsf{ff}&=&\mathsf{ff} \\
      \mathrm{strictOr}&\bot&\_&=&\bot \\
      \mathrm{strictOr}&\_&\bot&=&\bot \\
    \end{array}
    \qquad
    \begin{array}[t]{l@{(}l@{,~}l@{)~}c@{~}l}
      \mathrm{shortCircuitOR}&\mathsf{tt}&\_&=&\mathsf{tt} \\
      \mathsf{shortCircuitOR}&\mathsf{ff}&x&=&x \\
      \mathsf{shortCircuitOR}&\bot&\_& =& \bot
    \end{array}
    \qquad
    \raisebox{-15ex}{
      \begin{tikzpicture}
        \node (top) at (0,0) {$(\mathsf{tt},\mathsf{ff})$};
        \node [below of=top,xshift=-0.5cm] (oi) {$(\bot,\mathsf{ff})$};
        \node [below of=top,xshift=0.5cm] (io) {$(\mathsf{tt},\bot)$};
        \node [below of=oi,xshift=0.5cm] (oo) {$(\bot,\bot)$};
        \draw (top) -- (oi);
        \draw (top) -- (io);
        \draw (oi) -- (oo);
        \draw (io) -- (oo);
      \end{tikzpicture}
    }
  \end{displaymath}
  In the poset $\mathbb{B}_\bot^2$, a typical poset of approximations of a fully defined element is shown to the right. For $\mathrm{strictOr}$, any approximation that isn't the fully defined input is mapped to $\bot$, while $\mathrm{shortCircuitOr}$ maps the partially defined $(\mathsf{tt},\bot)$ to $\mathsf{tt}$. Thus, even though these functions operate identically on fully defined inputs, they differ in their \emph{derivatives} on partially defined input, exposing how . That they are cm can be checked by examining their restrictions' behaviour. If we take the approximations $(\bot,\mathsf{ff})$ and $(\mathsf{tt},\bot)$, then their meet is $(\bot,\bot)$; $\mathrm{strictOr}$ maps all three elements to $\bot$, so is cm here since $\bot \wedge \bot = \bot$; and $\mathrm{shortCircuitOR}$ has $(\bot,\mathsf{ff}) \mapsto \bot$ and $(\mathsf{tt},\bot) \mapsto \mathsf{tt}$, the meet of which is $\bot = \mathrm{shortCircuitOR}(\bot,\bot)$. Other combinations can be checked similarly.
\end{example}

### 2025-06-09 main/first-order.tex
Some minor improvements to families construction.
Gentler notation of syntactic identifiers.

The action of the families construction, regarded as a covariant 2-functor $\Cat \to \Cat$, sends functors $F:
\cat{C} \to \cat{D}$ to functors $\Fam(F): \Fam(\cat{C}) \to \Fam(\cat{D})$ that use $F$ to reassign the
target category by postcomposing with $F$, sending objects $(X, \partial X)$ to $(X, F \comp \partial X)$ and
morphisms $(f, \partial f)$ to $(f, F \comp \partial f)$, where $F \comp \partial f$ denotes the natural
transformation with $(F \comp \partial f)(x) = F(\partial f(x))$. If $F$ happens to be a (strong) monad on
$\cat{C}$, then $\Fam(F)$ is a (strong) monad on $\Fam(\cat{C})$, computed by lifting the original monad $F:
\cat{C} \to \cat{C}$ pointwise over families.

@@

For a first-order language with a primitive type $\syn{nat}$ and primitive operations such as $+$, the
``leaves'' of the approximation lattices would provided by a model of $\syn{nat}$ and its operations in
$\Fam(\LatGal)$, for example assigning $\syn{nat}$ to the object $(\Nat,n \mapsto \Two)$, where the family is
fully present ($\top$) or fully absent ($\bot$) from the approximation (as shown in the example in component
is constantly $\Two = (\bot,\top)$, the two-point lattice indicating whether a given natural number
\secref{introduction}). These ``leaf'' objects would then be combined by the interpretation of composite types
like sums and products, providing suitable derived approximation structures for composite values.

One might also like to allow the programmer to embed additional ``approximation points'' into composite
structures, by providing the language with a \emph{lifting} monad $\syn{L}$. Rather than the usual definition
of $\syn{list}\;\syn{nat}$ as $\mu\alpha.1 + \syn{nat} \times \alpha$, one could then define (within the
language) a type of ``approximation lists'' $\mu\alpha.\syn{L}\;(1 + \syn{L}\;\syn{nat} \times \alpha)$, along
with an approximating version of multiplication $*_{\text{appr}}: \syn{L}\;\syn{nat} \times \syn{L}\;\syn{nat}
\to \syn{L}\;\syn{nat}$ expressible as
\[x *_{\text{appr}} y = \tmBind{x}{\tmFun{x'}{\syn{if}\;x' =
0\;\syn{then}\;\tmReturn{0}\;\syn{else}\;\tmBind{y}{\tmFun{y'}{\tmReturn{x' * y'}}}}}\].

>>>>

The families construction, regarded as a covariant 2-functor $\Fam(-): \Cat \to \Cat$, sends functors $F:
\cat{C} \to \cat{D}$ to functors $\Fam(F): \Fam(\cat{C}) \to \Fam(\cat{D})$ that reassign the target category
by postcomposing with $F$, sending objects $(X, \partial X)$ to $(X, F \comp \partial X)$ and morphisms $(f,
\partial f)$ to $(f, F \comp \partial f)$, where $F \comp \partial f$ denotes the natural transformation where
$(F \comp \partial f)(x) = F(\partial f(x))$. If $F$ happens to be a (strong) monad on $\cat{C}$, then
$\Fam(F)$ is a (strong) monad on $\Fam(\cat{C})$, computed by lifting the original monad $F: \cat{C} \to
\cat{C}$ pointwise over families.

@@

For a first-order language with a primitive type $\syn{nat}$ and primitive operations such as $+$, the
``leaves'' of the approximation lattices would provided by a model of $\syn{nat}$ and its operations in
$\Fam(\LatGal)$. For example, we might assign $\syn{nat}$ to the object $(\Nat,n \mapsto \{n, \bot\})$, where
the family component is the two-point lattice indicating whether a given natural number is fully present ($n$)
or fully absent ($\bot$) from the approximation, as shown in the example in \secref{introduction}, and we
might then assign $+$ to the morphism $(\hat+,\partial(\hat+))$, with $\partial(\hat+)$ defined as
\begin{displaymath}
    \begin{array}[t]{l@{(}l@{,~}l@{)~}c@{~}l}
      \partial(\hat+)&n&m&=&n \mathbin{\hat+} m \\
      \partial(\hat+)&\bot&\_&=&\bot \\
      \partial(\hat+)&\_&\bot&=&\bot
    \end{array}
\end{displaymath}
These elementary approximations would then be combined by the interpretation of composite types like sums and
products, providing suitable derived approximation structures for composite values.

One might also like to allow the programmer to embed additional ``approximation points'' into composite
structures, by providing them with a \emph{lifting} monad $\synVar{L}$. Rather than the usual definition of
the type of lists of natural numbers as $\mu\alpha.1 + \synVar{nat} \times \alpha$, one could then define
(within the language) a type of ``approximation lists'' $\mu\alpha.\synVar{L}\,(1 + \synVar{L}\,\synVar{nat}
\times \alpha)$, along with an approximating version of multiplication $*_{\text{appr}}:
\synVar{L}\,\synVar{nat} \times \synVar{L}\,\synVar{nat} \to \synVar{L}\,\synVar{nat}$ given as as
\[x *_{\text{appr}} y = \tmBind{x}{\tmFun{x'}{\syn{if}\;x' =
0\;\syn{then}\;\tmReturn{0}\;\syn{else}\;\tmBind{y}{\tmFun{y'}{\tmReturn{x' * y'}}}}}\].

\noindent This would allow the programmer to emulate the short-circuiting behaviour of $\mathrm{strictOr}$ in
\exref{strict-short-circuit}, or the short-circuiting multiplication used in \citet{perera22} for slicing
matrix convolutions, without baking this specific approximating behaviour into the interpretation of
individual operators.
